Constructing ordinal outcomes for clinical trials

### 2025-06-19 constructing_ordinal_outcomes.tex
Many updates; start to restructure initial sections

\begin{abstract}
  Ordinal outcomes or endpoints are an increasingly popular choice in
  clinical trials.
  They have been recently popular in the study of COVID-19 and in
  adaptive platform
  trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes, aimed at clinicians and their statistical collaborators.
  We categorize and illustrate ways combining existing binary,
  ordinal, or continuous
  outcomes through three broad mechanisms: sequential composition
  (overriding), hierarchical composition (tie-breaking), and parallel
  composition.
  We discuss how this fine-graining impacts validity and efficiency, and also
  suggest reasons why coarser outcomes might be preferred.
  We also situate ``generalized pairwise comparisons'' and
  ``prioritized outcomes'' approaches, like the win
  ratio and desirability of outcome
  ranking (DOOR) methodologies, in the broader context of ordinal
  outcomes, suggesting that methods and approaches from each field
  might be useful in the other.
\end{abstract}

@@

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
of drugs repurposed to treat the disease \todo{cite}. They have also
been a popular choice in adaptive platform trials \todo{cite}.

Ordinal outcomes rank patient outcomes from better to worse, but do
not compare \emph{differences} in patient outcomes. Thus binary
outcomes are ordinal, and so are continuous outcomes (though
``treating them ordinally'' means ignoring numeric values).

Some advantages of ordinal outcomes are that they can
\begin{enumerate}
  \item Provide more information than binary outcomes, making study
    designs more sensitive and precise, and
    therefore more efficient and ethical

  \item Allow for the incorporation of intercurrent events into one's primary
    outcome, avoiding complexities of alternative estimand strategies like
    principal stratum estimands (particular for truncating events
    like death) while
    yielding more statistical efficiency than traditional binary
    composite outcomes

@@

  \item Allow for the combination of discrete and continous outcome
    measures (for
      example a numerical quality of life scale with death as the
      overriding worst
    outcome) and allow for the combination of longitudinal and timepoint or
    ``state'' outcomes in a single measure.

\end{enumerate}

@@

We want outcomes that
\begin{enumerate}
  \item are \textbf{valid} in that they reflect patient benefit and harm:
    better/worse outcomes are judged respectively better/worse by patients, are
    objectively respectively better/worse for patients, or are reliable
    surrogates/predictors of outcomes that are respectively better/worse for
    patients
  \item are \textbf{efficient} in the sense of capturing
    sufficient information to
    discriminate among better- and worse-off patients without large
    sample sizes.
\end{enumerate}
(For the purposes of this article, set aside ease of collection, low
missingess, objective ascertainability, etc.)

@@

@@

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{\textbf{Hierarchical composition (complete tie-breaking)
      illustrated with the
    the ORBITA-2 outcome.} ``Angina
    Symptom Score'' ordinal outocome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) hierarchical composition or complete tie-breaking (of units
      of antianginal medication by
    episodes of angina) and also (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

@@

\section{Conclusions}

>>>>

\begin{abstract}
  Ordinal outcomes (or ``endpoints'') are an increasingly popular choice in
  clinical trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes, aimed at clinicians and their statistical collaborators.
  We categorize and illustrate ways of combining existing binary,
  ordinal, or continuous
  outcomes through three broad mechanisms: sequential composition
  (overriding), hierarchical composition (tie-breaking), and parallel
  composition.
  We discuss how this fine-graining impacts validity and efficiency, and also
  suggest reasons why coarser outcomes might be preferred.
  We also situate ``generalized pairwise comparisons'' and
  ``prioritized outcomes'' approaches, like the win
  ratio and desirability of outcome
  ranking (DOOR) methodologies, in the broader context of ordinal
  outcomes, suggesting that methods and approaches from each field
  might be useful in the other.
\end{abstract}

@@

\subsection{Ordinal outcomes and their advantages}

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
\citep{beigelRemdesivirTreatmentCovid192020,
marshallMinimalCommonOutcome2020}. They have also
been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.

Ordinal outcomes rank patient outcomes from better to worse, but
\emph{differences} in patient outcomes are
not treated as meaningful. Binary
outcomes are ordinal -- they are the simplest kind of
ordinal outcome -- and so too are continuous outcomes, though
``treating them ordinally'' involves ignoring their numeric values.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}.

\textbf{Valid} outcomes reflect
patient benefit and harm: better/worse outcomes are judged respectively
better/worse by patients, are objectively respectively better/worse for
patients, or are reliable surrogates/predictors of outcomes that are
respectively better/worse for patients.

\begin{itemize}
  \item Allow for the incorporation of intercurrent events into one's primary
    outcome, avoiding complexities of alternative strategies like
    principal stratum
    estimands (particular for truncating events like death) while yielding more
    statistical efficiency than traditional binary composite outcomes

@@

  \item Allow for the combination of discrete and continous outcome
    measures (for
      example a numerical quality of life scale with death as the
      overriding worst
    outcome) and allow for the combination of longitudinal and timepoint or
    ``state'' outcomes in a single measure.
\end{itemize}

\textbf{Efficient} outcomes
capture sufficient information to discriminate among better- and worse-off
patients without large sample sizes.

\begin{itemize}
  \item Provide more information than binary outcomes, making study
    designs more sensitive and precise, and
    therefore more efficient and ethical
\end{itemize}

For the purposes of this
article, set aside ease of collection, low missingess, objective
ascertainability, etc.

A key feature of ordinal outcomes is the ease with which they can be
\emph{combined}.

@@

@@

\subsection{Objectives/contributions}

@@

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{\textbf{Hierarchical composition (complete tie-breaking)
      illustrated with the
    the ORBITA-2 outcome.} ``Angina
    Symptom Score'' ordinal outcome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) hierarchical composition or complete tie-breaking (of units
      of antianginal medication by
    episodes of angina) and also (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

@@

\section{Discussion and Conclusions}
